Properties of winning Iterated Prisoner's Dilemma strategies.

### 2020-03-30 paper/main.tex
minor tweaks I missed on the pr

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades: from the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated learning
structures such as neural networks, many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments, casting doubt on the generalizability
of conclusions. We analyze a large collection of \numberofstrategies typically known
strategies in \numberofalltournaments tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios.
Nevertheless there are several properties that heavily influence the best performing
strategies, refining the properties described by R. Axelrod in light of
recent and more diverse opponent populations. These are: be nice, be provocable and contrite,
be a little envious, be clever, and adapt to the environment, which includes the parameters
of the tournament (e.g. noise) and the population of opponents. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities,
or a proportion thereof in the case of noisy and probabilistically ending tournaments,
and that the manner in which a strategy achieves the ideal cooperation rate is crucial.
The features of high performing strategies reveal why strategies such as Tit For Tat
performed historically well in tournaments and why Zero Determinant strategies
typically do not fare well in tournament settings.
\end{abstract}

@@

In probabilistic ending tournaments with \(p_e \in [0, 1]\) the top ranks are
mostly occupied by defecting strategies such as Better and Better, Gradual
Killer, Hard Prober (all from~\cite{axelrodproject}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector, and a series of strategies based on finite
state automata introduced by Daniel Ashlock and Wendy Ashlock; Fortress 3,
Fortress 4 (both introduced in~\cite{Ashlock2006}), Raider~\cite{Ashlock2014}
and Solution B1~\cite{Ashlock2014}. The success of defecting strategies in
probabilistic ending tournaments is due to larger values of
\(p_e\) which lead to shorter matches (the expected number of rounds is $1\p_e$), so the
impact of the PD being iterated is subdued. This is captured by the Folk
Theorem~\cite{Fudenberg2009} as defecting strategies do better when the likelihood
of the game ending in the next turn increases.
This is demonstrated by Figure~\ref{fig:effect_of_probend}, which gives the
distributions of \(r\) for the top 6 strategies in probabilistic ending tournaments
over \(p_e\).

@@

Moreover, the manner in which a strategy achieves a given cooperation rate
relative to the tournament population average is important. Playing a strategy
that randomly cooperates with $C_mean$ is unlikely to be as effective as it will
randomly have cooperations matched with defections. In contrast, TFT naturally
achieves a cooperation rate near $C_mean$ by virtue of copying its opponent's
last move while also minimizing instances where it is exploited by an opponent
(cooperating while the opponent defects), at least in non-noisy tournaments.
\footnote{This also explains why Tit For N Tats does not fare well -- it fails
to achieve the proper cooperation ratio.}
TFT forces the opponent to pay back the (C, D) round with a (D, C) round before
returning to mutual cooperation. This explains why TFT performed well
in Axelrod's original tournaments as most strategies submitted to those tournaments
were typically cooperative and relatively few strategies used an easily exploitable
pattern. TFT does not appear in the top ranks in
these tournaments because it is too nice. Strategies like Grudger will always
defect after a fixed number of opponent defections, which allows them to
effectively exploit strategies like Alternator or stochastic strategies that
have a non-zero chance of cooperating after mutual defection, which TFT will
not do. Moreover in a noisy environment these strategies will naturally tend
toward always defecting, leading them to exploit strategies like Cooperator. In
such a scenario the noisy environment effectively voids Axelrod's rule to be nice,
allowing strategies to attempt exploitation, whereas in a noise-free environment,
exploitation is risky because several strategies exhibit a Grudger-like behavior,
reducing the overall value in attempting to exploit strategies like Cooperator.

>>>>

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades: from the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated learning
structures such as neural networks, many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a very small number of tournaments, casting doubt on the generalizability
of conclusions. We analyze a large collection of \numberofstrategies typically known
strategies in \numberofalltournaments tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios.
Nevertheless there are several properties that heavily influence the best performing
strategies, refining the properties described by R. Axelrod in light of
recent and more diverse opponent populations. These are: be nice, be provocable and contrite,
be a little envious, be clever, and adapt to the environment, which includes the parameters
of the tournament (e.g. noise) and the population of opponents. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities,
or a proportion thereof in the case of noisy and probabilistically ending tournaments,
and that the manner in which a strategy achieves the ideal cooperation rate is crucial.
The features of high performing strategies reveal why strategies such as Tit For Tat
performed historically well in tournaments and why zero-determinant strategies
typically do not fare well in tournament settings.
\end{abstract}

@@

In probabilistic ending tournaments with \(p_e \in [0, 1]\) the top ranks are
mostly occupied by defecting strategies such as Better and Better, Gradual
Killer, Hard Prober (all from~\cite{axelrodproject}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector, and a series of strategies based on finite
state automata introduced by Daniel Ashlock and Wendy Ashlock; Fortress 3,
Fortress 4 (both introduced in~\cite{Ashlock2006}), Raider~\cite{Ashlock2014}
and Solution B1~\cite{Ashlock2014}. The success of defecting strategies in
probabilistic ending tournaments is due to larger values of
\(p_e\) which lead to shorter matches (the expected number of rounds is \(1 / p_e\)), so the
impact of the PD being iterated is subdued. This is captured by the Folk
Theorem~\cite{Fudenberg2009} as defecting strategies do better when the likelihood
of the game ending in the next turn increases.
This is demonstrated by Figure~\ref{fig:effect_of_probend}, which gives the
distributions of \(r\) for the top 6 strategies in probabilistic ending tournaments
over \(p_e\).

@@

Moreover, the manner in which a strategy achieves a given cooperation rate
relative to the tournament population average is important. Playing a strategy
that randomly cooperates with $C_{\text{mean}}$ is unlikely to be as effective as it will
randomly have cooperations matched with defections. In contrast, TFT naturally
achieves a cooperation rate near $C_{\text{mean}}$ by virtue of copying its opponent's
last move while also minimizing instances where it is exploited by an opponent
(cooperating while the opponent defects), at least in non-noisy tournaments.
\footnote{This also explains why Tit For N Tats does not fare well -- it fails
to achieve the proper cooperation ratio.}
TFT forces the opponent to pay back the (C, D) round with a (D, C) round before
returning to mutual cooperation. This explains why TFT performed well
in Axelrod's original tournaments as most strategies submitted to those tournaments
were typically cooperative and relatively few strategies used an easily exploitable
pattern. TFT does not appear in the top ranks in
these tournaments because it is too nice. Strategies like Grudger will always
defect after a fixed number of opponent defections, which allows them to
effectively exploit strategies like Alternator or stochastic strategies that
have a non-zero chance of cooperating after mutual defection, which TFT will
not do. Moreover in a noisy environment these strategies will naturally tend
toward always defecting, leading them to exploit strategies like Cooperator. In
such a scenario the noisy environment effectively voids Axelrod's rule to be nice,
allowing strategies to attempt exploitation, whereas in a noise-free environment,
exploitation is risky because several strategies exhibit a Grudger-like behavior,
reducing the overall value in attempting to exploit strategies like Cooperator.
